Normalization: A pipeline to analyse time-course gene expression data

### 2019-02-23 scripts/01_quality_control_normalization.Rmd
FIX More text + small fixes in QC + normalization

## Pre-processing

The micro-array data were obtained from NCBI Gene Expression Omnibus (GEO),
with accession number GSE63786. The dataset has a total of 209 samples, with
45018 probes.

@@

```{r reading_targets, results="hide", fig.width=8, fig.height=8}
# Read in the data using limma.
targets = limma::readTargets()
RG = limma::read.maimages(
  targets, source="agilent",
  path=data_dir,
  other.columns="gIsWellAboveBG",
  green.only=TRUE)

# Preform background correction using norm-exponential strategy and normalize
# across samples using quantile normalization.
RG = backgroundCorrect(RG, method="normexp")
RG.q = normalizeBetweenArrays(RG, method="quantile")

# Plot the densities of the probes before and after normalization.
par(mfrow=c(1, 2))
plotDensities(RG, col="black", legend=FALSE)
plotDensities(RG.q, col="black", legend=FALSE)
```

@@

```{r order_genes}
# Reorder genes on condition, time, and replicate

library(ggfortify)
pca_data = prcomp(t(data), rank=5, center=TRUE, scale=TRUE) 
percent_var = round(100 * attr(pca_data, "percentVar"))

meta["Time"] = as.factor(unlist(meta["Time"]))
meta["Replicate"] = as.factor(unlist(meta["Replicate"]))
par(mfrow=c(1, 2))
autoplot(pca_data, data=meta, colour="Group", shape="Replicate")
autoplot(pca_data, data=meta, colour="Time", shape="Replicate")
```

We then plot the correlation matrix

```{r fig.width=6, fig.height=6}
library(NMF)

# Reorder genes on condition, time, and replicate
ord = order(
  meta$Group,
  meta$Time,
  meta$Replicate)
data = data[, ord]
meta = meta[ord, ]

data_corr = cor(data, method="pearson")
aheatmap(data_corr,
         Colv=NA, Rowv=NA,
         annCol=meta[,
		     c("Group", "Time",
			"Replicate")],
         annLegend=FALSE, 
	 annColors=ann_colors,
         main="Correlation plot after Normalization")
```

>>>>

## Pre-processing and normalization

The micro-array data were obtained from NCBI Gene Expression Omnibus (GEO),
with accession number GSE63786. The dataset has a total of 209 samples, with
45018 probes (1 sample was removed by the authors prior to uploading the
dataset on GEo).

@@

```{r reading_targets, results="hide", fig.width=8, fig.height=4}
# Read in the data using limma.
targets = limma::readTargets()
RG = limma::read.maimages(
  targets, source="agilent",
  path=data_dir,
  other.columns="gIsWellAboveBG",
  green.only=TRUE)

# Preform background correction using norm-exponential strategy and normalize
# across samples using quantile normalization.
RG = backgroundCorrect(RG, method="normexp")
RG.q = normalizeBetweenArrays(RG, method="quantile")
```

Plotting the smoothed empirical densities of the probe values before
normalization shows considerable variations betweens arrays. Once the data has
been normalized across array using quantile normalization, the smoothed
empirical densities of each sample perfectly overlap.

```{r plotting_densities, results="hide", fig.width=8, fig.height=4}
# Plot the densities of the probes before and after normalization.
par(mfrow=c(1, 2))
plotDensities(RG, col="black", legend=FALSE)
plotDensities(RG.q, col="black", legend=FALSE)
```

@@

## Exploratory analysis and quality control

Plotting the smoothed empirical densities of probe values for each sample and
estimating the variations across samples before and after normalization is a
first quality control. Typically, two other quality control and exploratory
analysis steps are also performed before and after normalization: (1) low
dimensionality embedding of the samples; (2) correlation plots between each
samples. In both cases, we expect a strong biological signal, while replicate
samples should be strongly clustered or correlated with one another.

Let us first perform the PCA analysis. Here, we perform a PCA of rank 3 of the
centered and scaled gene expression data.

```{r pca, fig.width=8, fig.height=8}
# Reorder genes on condition, time, and replicate

library(ggfortify)
pca_data = prcomp(t(data), rank=3, center=TRUE, scale=TRUE) 
percent_var = round(100 * attr(pca_data, "percentVar"))
```

We then plot the two first components, and color each sample by (1) its
condition; (2) its sampling time. We use different markers for each replicate.
We also plot the second and third components in the second row.

```{r pca_plots, fig.width=8, fig.height=8}
# FIXME Add legend
par(mfrow=c(2, 2))
plot(pca_data$x[, "PC2"], pca_data$x[, "PC1"],
     col=ann_colors$Group[meta$Group],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
     xlab="PC2", ylab="PC1")
plot(pca_data$x[, "PC2"], pca_data$x[, "PC1"],
     col=ann_colors$Time[as.factor(meta$Time)],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
     xlab="PC2", ylab="PC1")

plot(pca_data$x[, "PC2"], pca_data$x[, "PC3"],
     col=ann_colors$Group[meta$Group],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
     xlab="PC2", ylab="PC3")
plot(pca_data$x[, "PC2"], pca_data$x[, "PC3"],
     col=ann_colors$Time[as.factor(meta$Time)],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
     xlab="PC2", ylab="PC3")
```

Then, we plot the pearson correlation across each samples. We order the
samples by their Group (treatment) and Time (the time of sampling).
Additionally, in this example, we order each treatment by strength of the
pathogenicity of the treatment: Control, Kawasaki, California, low dose of
Vietnam, then high dose of Vietnam.

```{r  correlation_plot, fig.width=6, fig.height=6}
library(NMF)

# Reorder genes on condition, time, and replicate
meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 1, 2, 5, 4)])
ord = order(
  meta$Group,
  meta$Time,
  meta$Replicate)
data = data[, ord]
meta = meta[ord, ]

data_corr = cor(data, method="pearson")

aheatmap(data_corr,
         Colv=NA, Rowv=NA,
         annCol=meta[,
		     c("Group", "Time",
		       "Replicate")],
         annLegend=FALSE, 
	 annColors=ann_colors,
         main="Correlation plot")
```

We can already see interesting patterns: the cross-correlation amongst samples
taking from the untreated mice is higher than the cross correlation amongst
the rest of the treatments.
